---
title: "Supplementary Information for Diverse lithic production strategies in southwest China during Late Middle Pleistocene"
---

<!-- This is the format for text comments that will be ignored during renderings. Do not put R code in these comments because it will not be ignored. -->

<!-- With the following code you can access and display values from the yml header above. -->

Keywords: `r rmarkdown::metadata$keywords`

Highlights: `r rmarkdown::metadata$highlights`

<!-- The following code chunk defines some general settings how code chunks should behave. -->

```{r setup, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = TRUE,
  comment = "#>",
  fig.path = "../figures/",
  dpi = 300
)
```

<!-- The actual document text starts here: -->

Read in the data.

```{r get-data}
# Note the path that we need to use to access our data files when rendering this document
library(readxl)
library(tidyverse)

file_name <-
  here::here("analysis/data/raw_data/artefacts-after Sam_Dec2018_quina_modified-2019.xls")

flakes <- as.data.frame(read_excel(file_name, sheet = "flake basics"))
cores <- as.data.frame(read_excel(file_name, sheet = "core basics"))
debris <- as.data.frame(read_excel(file_name, sheet = "chunk&debris"))
retouch <- as.data.frame(read_excel(file_name, sheet = "retouch"))
core_flake_scars_1 <- as.data.frame(read_excel(file_name, sheet = "core flake scars"))
edge_angle_1 <- as.data.frame(read_excel(file_name, sheet = "edge angle"))
retouched_pebble_chunk <- as.data.frame(read_excel(file_name, sheet = "Pebble&chunk"))

scar_dir <- as.data.frame(read_excel(file_name, sheet = 'scar direction'))

core_flake_scars <- core_flake_scars_1[core_flake_scars_1$number %in% cores$number, ]

edge_angle <- edge_angle_1[edge_angle_1$number %in% retouch$number, ]
```

Prepare and clean the data

```{r}
# prepare the data --------------------------------------------------

require(Ckmeans.1d.dp)
result <- Ckmeans.1d.dp(flakes$mass[!is.na(flakes$mass)])

n_clusters <- length(result$size)
n_clusters_30 <- length(result$size[result$size > 30])

# assign cluster ID to table
flakes_clustered <-
  flakes %>%
  filter(!is.na(mass)) %>%
  mutate(cluster = result$cluster) %>%
  right_join(flakes) %>%
  filter(cluster %in% 1:n_clusters_30)

# clean data to help with analysis
flakes$retouched <- ifelse(grepl("ret", flakes$type1), "retouched", "unretouched")
flakes <- flakes[!flakes$material %in% c('sandstone', 'quartz'),]
```

Separate Levallois and Non-Levallois flakes for comparison

```{r}
library(tidyverse)
library(ggbeeswarm)
library(papaja)

# clean data to help with analysis
flakes$retouched <- ifelse(grepl("ret", flakes$type1), "retouched", "unretouched")
flakes <-
  flakes[!flakes$material %in% c('sandstone', 'quartz'), ] %>%
  mutate(complet_flk = ifelse(
    type1 %in% c(
      "flake",
      "blade",
      "KBW",
      "tablet",
      "leva",
      "leva?",
      "DBD",
      "crest"
    ),
    "complete flake",
    ifelse(
      type1 %in% c("flake brk", "blade brk", "LVFB", "proximal"),
      "flake breaks",
      "other"
    )
  )) 

# how many levallois?
# unique(flakes$type1)

leva <- flakes %>% 
  filter(grepl("LVT|LVF|LVFB", type1)) %>%
  filter(!grepl("LVF_", type1))

non_leva <- flakes %>% 
  filter(!grepl("LVT|LVF|LVFB", type1))

flakes_L <-  
  mutate(flakes, L = ifelse(grepl("LVT|LVF|LVFB", type1), 
                            "Levallois", "Non-Levallois"))  
```

Results of descriptive statistics for Levallois and non-Levallois flakes. ‘PLF’ stands for preferential Levallois flake; ‘CF’ stands for complete flake. This is table 1 in the main text

```{r}
flakes_complete_non_Lev <- 
  flakes_L %>% # only complete flakes for non-Levallois
  filter(complet_flk == "complete flake")

flakes_Lev  <- 
  flakes_L %>% 
  filter(L == "Levallois")

# combine both here
flakes_Lev_non_Lev <- 
bind_rows(flakes_complete_non_Lev,
          flakes_Lev) %>% 
  as_tibble()
  

# check the CVs
CV <- function(the_vector){
  mean_ <- mean(the_vector, na.rm = TRUE)
  sd_ <- sd(the_vector, na.rm = TRUE)
  cv <- (sd_/mean_) * 100
  round(cv, 3)
}

# prepare mean and sd functions
mean_t <- function(x) mean(x, na.rm = TRUE)
sd_t <- function(x) sd(x, na.rm = TRUE)

# for the table 'Results of descriptive statistics for Levallois and non-Levallois flakes. ‘PLF’ stands for preferential Levallois flake; ‘CF’ stands for complete flake.'
compare_lev_non_lev_vars_tbl <- 
flakes_Lev_non_Lev %>% 
  as_tibble() %>% 
  select(L, where(is.numeric)) %>% 
  group_by(L) %>% 
  summarise(across(where(is.numeric),
            list(mean = mean_t,
                 sd = sd_t,
                 cv = CV),
            .names = "{.fn}.{.col}")) %>%
  pivot_longer(-L) %>%
  separate(col = name,
           into = c("stat", "var"),
           sep = "\\.") %>% 
  mutate(L = ifelse(L == "Levallois",
                    "PLF",
                    "CF")) %>% 
  unite("lev_stat", c(stat, L), 
        sep = "_") %>% 
  pivot_wider(names_from = lev_stat ) %>% 
  mutate(difference_in_cv = cv_PLF - `cv_CF`) %>% 
  mutate(across(where(is.numeric), round, 2))

library(gt) 
tab <- 
  compare_lev_non_lev_vars_tbl %>%
  gt(rowname_col = "var") %>% 
  tab_spanner(
    label = "Mean (mm)",
    columns = vars(mean_PLF, 
                   mean_CF)) %>% 
  tab_spanner(
    label = "SD",
    columns = vars(sd_PLF, 
                   sd_CF)  
  ) %>% 
  tab_spanner(
    label = "CV (%)",
    columns = vars(cv_PLF, 
                   cv_CF,
                   difference_in_cv) 
  )

tab

```

Are CV values lower for Levallois flakes than for non-Levallois flakes? This result is presented in line in the text in the main text.

```{r}
compare_lev_non_lev_vars_tbl  %>% 
  select(contains("cv_")) %>% 
  stack() %>% 
  wilcox.test(values ~ ind, 
              data = .) %>% 
  apa_print() %>% 
  .$statistic %>% 
  str_remove_all("\\$|\\{|\\}|\\\\|mathit")

```


## Comparison of Levallois flakes vs non-Levallois flakes

Are the levallois pieces bigger by mass? 

```{r}
# plotting setting
point_alpha <- 0.1

# is it significant?
t_test_str_l_mass <- 
apa_print(t.test(data = flakes_L, mass ~ L))$statistic %>% 
  str_remove_all("\\$|\\{|\\}|\\\\|mathit")

# plot
compare_lev_non_lev_mass_plot <- 
ggplot(flakes_L, aes(x = L, 
                     y = mass)) +
  geom_boxplot() +
  geom_quasirandom(alpha = point_alpha) +
  scale_y_log10() + 
  theme_bw() + 
  xlab("") +
  ylab("Mass (g)") +
  ggtitle(paste0("Flake Mass: ", t_test_str_l_mass))

compare_lev_non_lev_mass_plot
```

Are the levallois pieces bigger by max dimension?

```{r}
# is it significant?
t_test_str_l_maxdim <- 
apa_print(t.test(data = flakes_L, 
                 max_dimension ~ L))$statistic %>% 
  str_remove_all("\\$|\\{|\\}|\\\\|mathit")

# plot
compare_lev_non_lev_maxdim_plot <- 
ggplot(flakes_L, aes(x = L, 
                     y = max_dimension)) +
  geom_boxplot() +
  geom_quasirandom(alpha = point_alpha) +
  scale_y_log10() + 
  theme_bw() + 
  xlab("") +
  ylab("Maximum Dimension (mm)") +
  ggtitle(paste0("Flake Max. Dim.: ", t_test_str_l_maxdim))

compare_lev_non_lev_maxdim_plot
```

Are the levallois pieces bigger by thickness?

```{r}
# is it significant?
t_test_str_l_thick <- 
apa_print(t.test(data = flakes_L,
                 `Thickness_50%max` ~ L))$statistic %>% 
  str_remove_all("\\$|\\{|\\}|\\\\|mathit")

# plot
compare_lev_non_lev_thick_plot <- 
ggplot(flakes_L, aes(x = L, 
                     y = `Thickness_50%max`)) +
  geom_boxplot() +
  geom_quasirandom(alpha = point_alpha) +
  scale_y_log10() + 
  theme_bw() + 
  xlab("") +
  ylab("Thickness (mm)") +
  ggtitle(paste0("Flake Thickness: ", t_test_str_l_thick))

compare_lev_non_lev_thick_plot
```

## Comparison of Quina tools vs non-Quina tools

Do the Quina tools have different edge angles?

```{r}
library(reshape2)
edge_angle_2 <- edge_angle

edge_angle_2[, 2:(ncol(edge_angle_2)-1)] <- atan(edge_angle[, 2:(ncol(edge_angle_2)-1)]/2/3)/pi*180*2

df_angle <- melt(edge_angle_2[, 1:(ncol(edge_angle_2)-1)])

df_angle$variable <- substr(df_angle$variable, 1, 9)

colnames(df_angle)[2:3] <- c("section", "angle")

quina_id <- flakes$number[which(flakes$type3 == "quina")]

df_angle$quina <- ifelse(df_angle$number %in% quina_id, "Quina", "Non-Quina")

# siginificance test
t_test_str_q_angle <- 
df_angle %>% 
  drop_na("angle") %>% 
  t.test(data = ., angle ~ quina) %>% 
  apa_print %>% 
  .$statistic %>% 
  str_remove_all("\\$|\\{|\\}|\\\\|mathit")

compare_qui_non_qui_angle_plot <-
  ggplot(df_angle,
         aes(y = angle,
             x = quina)) +
  geom_boxplot() +
  geom_quasirandom(alpha = point_alpha) +
  theme_bw() +
  ylab("Edge angle (\u00B0)") +
  xlab("") +
  ggtitle(paste0("Flake Edge Angle: ", t_test_str_q_angle) )

compare_qui_non_qui_angle_plot

```

Do the Quina tools have different thickness?

```{r}
df_thick <- melt(flakes[, c("number", "Thickness_25%max", "Thickness_50%max", "Thickness_75%max")])

df_thick$quina <- ifelse(df_thick$number %in% quina_id, "Quina", "Non-Quina")

df_thick$label <- substring(df_thick$variable, 11, 13)

df_thick_tt_test <- 
df_thick %>% 
  nest_by(label) %>% 
  mutate(t_test = list(
    t.test(data = data, value ~ quina) %>% 
      apa_print %>% 
      .$statistic %>% 
  str_remove_all("\\$|\\{|\\}|\\\\|mathit")
  ))

compare_qui_non_qui_thick_plot <-
  ggplot(df_thick,
         aes(y = value,
             x = quina)) +
  geom_boxplot() +
  geom_quasirandom(alpha = 0.05) +
  geom_text(data = df_thick_tt_test,
            aes(label = t_test,
                x = 1.5, 
                y = 60)) +
  theme_bw() +
  facet_wrap(~label) +
  ylab("Thickness (mm)") + 
  xlab("") 

compare_qui_non_qui_thick_plot

```

Do the Quina tools have different GIUR?

```{r}
# plot GIUR on Quina  --------------------------------------------------

## calculate GIUR
# looking at the GIUR data... there are some duplicate artefact numbers, which is a nuisance

# make the number ID unique
retouch$number <- make.unique(as.character(retouch$number), sep = "_")

retouch_giur_per_section <-
  retouch %>%
  dplyr::select(grep("^number$|_t|_T", names(retouch))) %>%
  gather(variable, value, -number) %>%
  separate(variable, c("section", "tee"), sep = 10) %>%
  pivot_wider(names_from =  tee, values_from = value) %>%
  mutate(t_T = t / T)  %>%
  filter(t_T <= 1)

retouch_giur_per_artefact <-
  retouch_giur_per_section %>%
  group_by(number) %>%
  summarise(mean_giur = mean(t_T, na.rm = TRUE)) %>% 
  mutate(quina = ifelse(number %in% quina_id, "Quina", "Non-Quina"))

df_giur_clustered_tt_test <- 
retouch_giur_per_artefact %>% 
    t.test(data = ., 
           mean_giur ~ quina) %>% 
      apa_print %>% 
      .$statistic %>% 
  str_remove_all("\\$|\\{|\\}|\\\\|mathit")

# code for raincloud plot
source("https://gist.githubusercontent.com/benmarwick/2a1bb0133ff568cbe28d/raw/fb53bd97121f7f9ce947837ef1a4c65a73bffb3f/geom_flat_violin.R")

compare_qui_non_qui_giur_plot <-
  ggplot(retouch_giur_per_artefact) +
  aes(y = mean_giur,
      x = quina) +
  geom_flat_violin(position = position_nudge(x = .25, 
                                             y = 0), 
                   alpha = .8,
                   adjust = 0.35)  +
  geom_quasirandom( 
             size = 1.5, 
             width = 0.25,
             alpha = point_alpha) +
  geom_boxplot(
        width = .25, 
               outlier.shape = NA, 
               alpha = 0.25
               ) +
  ggtitle(paste0("Flake GIUR: ", df_giur_clustered_tt_test)) +
  coord_flip() +
  theme_bw() +
  ylab("GIUR") +
  xlab("")

compare_qui_non_qui_giur_plot

```

Plot combining multiple panels, this is Figure 1 is the main text

```{r}
## plot figures together.
library(cowplot)
figure_leva_quina <-
  cowplot::plot_grid(plot_grid(compare_lev_non_lev_mass_plot, 
                               compare_lev_non_lev_maxdim_plot, 
                               compare_lev_non_lev_thick_plot ,
                               labels = "auto",
                               label_size = 12,
                               ncol = 3),
                     plot_grid( 
                               compare_qui_non_qui_thick_plot ,
                               labels = c("d"),
                               label_size = 12,
                               ncol = 1,
                               rel_widths = c(1,2)),
                     plot_grid(compare_qui_non_qui_angle_plot,
                               compare_qui_non_qui_giur_plot , 
                               labels= c("e", "f"),
                               ncol = 2,
                               #rel_widths = c(1.5, 1),
                               label_size = 12),
                     nrow = 3,
                     align = 'v',
                     axis = 'lr',
                     vjust = 1.5,
                     hjust = -0.5,
                     scale = 0.9) #

save_plot(here::here("analysis/figures/fig-1-Figures-for-Leva-and-quina.png"),
                     figure_leva_quina,
          base_aspect_ratio = 1,
          base_height = 10, #18,
          base_width =  14) #22)

figure_leva_quina
```


\newpage

### Colophon

This report was generated on `r Sys.time()` using the following computational environment and dependencies: 

```{r colophon, cache = FALSE}
# which R packages and versions?
if ("devtools" %in% installed.packages()) devtools::session_info()
```

The current Git commit details are:

```{r}
# what commit is this file at? 
if ("git2r" %in% installed.packages() & git2r::in_repository(path = ".")) git2r::repository(here::here())  
```
