---
title: "Supplementary Information for Diverse lithic production strategies in southwest China during Late Middle Pleistocene"
author:
  - Myfirstname Mylastname:
      email: fl@oneplace.edu
      institute: [UofO]
      correspondence: true
  - Yourfirstname Yourlastname:
      email: fl@another.edu
      institute: [UofA]
      correspondence: false
institute:
  - UofO: University of One Place
  - UofA: University of Another Place
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  Text of abstract
keywords: |
  keyword 1; keyword 2; keyword 3
highlights: |
  These are the highlights. 
---

<!-- This is the format for text comments that will be ignored during renderings. Do not put R code in these comments because it will not be ignored. -->

<!-- With the following code you can access and display values from the yml header above. -->

Keywords: `r rmarkdown::metadata$keywords`

Highlights: `r rmarkdown::metadata$highlights`

<!-- The following code chunk defines some general settings how code chunks should behave. -->

```{r setup, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  comment = "#>",
  fig.path = "../figures/",
  dpi = 300
)
```

<!-- The actual document text starts here: -->

Read in the data.

```{r get-data}
# Note the path that we need to use to access our data files when rendering this document
library(readxl)

file_name <- here::here("analysis/data/raw_data/artefacts-after Sam_Dec2018_quina_modified.xls") 

flakes <- as.data.frame(read_excel(file_name, sheet = "flake basics"))
cores <- as.data.frame(read_excel(file_name, sheet = "core basics"))
debris <- as.data.frame(read_excel(file_name, sheet = "chunk&debris"))
retouch <- as.data.frame(read_excel(file_name, sheet = "retouch"))
core_flake_scars_1 <- as.data.frame(read_excel(file_name, sheet = "core flake scars"))
edge_angle_1 <- as.data.frame(read_excel(file_name, sheet = "edge angle"))
retouched_pebble_chunk <- as.data.frame(read_excel(file_name, sheet = "Pebble&chunk"))

scar_dir <- as.data.frame(read_excel(file_name, sheet = 'scar direction'))

core_flake_scars <- core_flake_scars_1[core_flake_scars_1$number %in% cores$number, ]

edge_angle <- edge_angle_1[edge_angle_1$number %in% retouch$number, ]
```

Prepare and clean the data

```{r}
# prepare the data --------------------------------------------------

require(Ckmeans.1d.dp)
result <- Ckmeans.1d.dp(flakes$mass[!is.na(flakes$mass)])

n_clusters <- length(result$size)
n_clusters_30 <- length(result$size[result$size > 30])

# assign cluster ID to table
flakes_clustered <-
  flakes %>%
  filter(!is.na(mass)) %>%
  mutate(cluster = result$cluster) %>%
  right_join(flakes) %>%
  filter(cluster %in% 1:n_clusters_30)

# clean data to help with analysis
flakes$retouched <- ifelse(grepl("ret", flakes$type1), "retouched", "unretouched")
flakes <- flakes[!flakes$material %in% c('sandstone', 'quartz'),]
```

Separate Levallois and Non-Levallois flakes for comparison

```{r}
library(tidyverse)
library(ggbeeswarm)
library(papaja)

# clean data to help with analysis
flakes$retouched <- ifelse(grepl("ret", flakes$type1), "retouched", "unretouched")
flakes <-
  flakes[!flakes$material %in% c('sandstone', 'quartz'), ] %>%
  mutate(complet_flk = ifelse(
    type1 %in% c(
      "flake",
      "blade",
      "KBW",
      "tablet",
      "leva",
      "leva?",
      "DBD",
      "crest"
    ),
    "complete flake",
    ifelse(
      type1 %in% c("flake brk", "blade brk", "LVFB", "proximal"),
      "flake breaks",
      "other"
    )
  )) 

# how many levallois?
# unique(flakes$type1)

leva <- flakes %>% 
  filter(grepl("LVT|LVF|LVFB", type1)) %>%
  filter(!grepl("LVF_", type1))

non_leva <- flakes %>% 
  filter(!grepl("LVT|LVF|LVFB", type1))

flakes_L <-  
  mutate(flakes, L = ifelse(grepl("LVT|LVF|LVFB", type1), 
                            "Levallois", "Non-Levallois"))  
```

Results of descriptive statistics for Levallois and non-Levallois flakes. ‘PLF’ stands for preferential Levallois flake; ‘CF’ stands for complete flake.

```{r}
flakes_complete_non_Lev <- 
  flakes_L %>% # only complete flakes for non-Levallois
  filter(complet_flk == "complete flake")

flakes_Lev  <- 
  flakes_L %>% 
  filter(L == "Levallois")

# combine both here
flakes_Lev_non_Lev <- 
bind_rows(flakes_complete_non_Lev,
          flakes_Lev) %>% 
  as_tibble()
  

# check the CVs
CV <- function(the_vector){
  mean_ <- mean(the_vector, na.rm = TRUE)
  sd_ <- sd(the_vector, na.rm = TRUE)
  cv <- (sd_/mean_) * 100
  round(cv, 3)
}

# prepare mean and sd functions
mean_t <- function(x) mean(x, na.rm = TRUE)
sd_t <- function(x) sd(x, na.rm = TRUE)

# for the table 'Results of descriptive statistics for Levallois and non-Levallois flakes. ‘PLF’ stands for preferential Levallois flake; ‘CF’ stands for complete flake.'
compare_lev_non_lev_vars_tbl <- 
flakes_Lev_non_Lev %>% 
  as_tibble() %>% 
  select(L, where(is.numeric)) %>% 
  group_by(L) %>% 
  summarise(across(where(is.numeric),
            list(mean = mean_t,
                 sd = sd_t,
                 cv = CV),
            .names = "{.fn}.{.col}")) %>%
  pivot_longer(-L) %>%
  separate(col = name,
           into = c("stat", "var"),
           sep = "\\.") %>% 
  mutate(L = ifelse(L == "Levallois",
                    "PLF",
                    "CF")) %>% 
  unite("lev_stat", c(stat, L), 
        sep = "_") %>% 
  pivot_wider(names_from = lev_stat ) %>% 
  mutate(difference_in_cv = cv_PLF - `cv_CF`) %>% 
  mutate(across(where(is.numeric), round, 2))

library(gt) 
tab <- 
  compare_lev_non_lev_vars_tbl %>%
  gt(rowname_col = "var") %>% 
  tab_spanner(
    label = "Mean (mm)",
    columns = vars(mean_PLF, 
                   mean_CF)) %>% 
  tab_spanner(
    label = "SD",
    columns = vars(sd_PLF, 
                   sd_CF)  
  ) %>% 
  tab_spanner(
    label = "CV (%)",
    columns = vars(cv_PLF, 
                   cv_CF,
                   difference_in_cv) 
  )

```

Are CV values lower for Levallois flakes than for non-Levallois flakes?

```{r}
compare_lev_non_lev_vars_tbl  %>% 
  select(contains("cv_")) %>% 
  stack() %>% 
  wilcox.test(values ~ ind, 
              data = .) %>% 
  apa_print() %>% 
  .$statistic %>% 
  str_remove_all("\\$|\\{|\\}|\\\\|mathit")

```


## Comparison of Levallois flakes vs non-Levallois flakes

Are the levallois pieces bigger by mass?

```{r}
# plotting setting
point_alpha <- 0.1

# is it significant?
t_test_str_l_mass <- 
apa_print(t.test(data = flakes_L, mass ~ L))$statistic %>% 
  str_remove_all("\\$|\\{|\\}|\\\\|mathit")

# plot
compare_lev_non_lev_mass_plot <- 
ggplot(flakes_L, aes(x = L, 
                     y = mass)) +
  geom_boxplot() +
  geom_quasirandom(alpha = point_alpha) +
  scale_y_log10() + 
  theme_bw() + 
  xlab("") +
  ylab("Mass (g)") +
  ggtitle(paste0("Flake Mass: ", t_test_str_l_mass))
```

Are the levallois pieces bigger by max dimension?

```{r}
# is it significant?
t_test_str_l_maxdim <- 
apa_print(t.test(data = flakes_L, 
                 max_dimension ~ L))$statistic %>% 
  str_remove_all("\\$|\\{|\\}|\\\\|mathit")

# plot
compare_lev_non_lev_maxdim_plot <- 
ggplot(flakes_L, aes(x = L, 
                     y = max_dimension)) +
  geom_boxplot() +
  geom_quasirandom(alpha = point_alpha) +
  scale_y_log10() + 
  theme_bw() + 
  xlab("") +
  ylab("Maximum Dimension (mm)") +
  ggtitle(paste0("Flake Max. Dim.: ", t_test_str_l_maxdim))
```

Are the levallois pieces bigger by thickness?

```{r}
# is it significant?
t_test_str_l_thick <- 
apa_print(t.test(data = flakes_L,
                 `Thickness_50%max` ~ L))$statistic %>% 
  str_remove_all("\\$|\\{|\\}|\\\\|mathit")

# plot
compare_lev_non_lev_thick_plot <- 
ggplot(flakes_L, aes(x = L, 
                     y = `Thickness_50%max`)) +
  geom_boxplot() +
  geom_quasirandom(alpha = point_alpha) +
  scale_y_log10() + 
  theme_bw() + 
  xlab("") +
  ylab("Thickness (mm)") +
  ggtitle(paste0("Flake Thickness: ", t_test_str_l_thick))
```

## Comparison of Quina tools vs non-Quina tools

Do the Quina tools have different edge angles?

```{r}
library(reshape2)
edge_angle_2 <- edge_angle

edge_angle_2[, 2:(ncol(edge_angle_2)-1)] <- atan(edge_angle[, 2:(ncol(edge_angle_2)-1)]/2/3)/pi*180*2

df_angle <- melt(edge_angle_2[, 1:(ncol(edge_angle_2)-1)])

df_angle$variable <- substr(df_angle$variable, 1, 9)

colnames(df_angle)[2:3] <- c("section", "angle")

quina_id <- flakes$number[which(flakes$type3 == "quina")]

df_angle$quina <- ifelse(df_angle$number %in% quina_id, "Quina", "Non-Quina")

# siginificance test
t_test_str_q_angle <- 
df_angle %>% 
  drop_na("angle") %>% 
  t.test(data = ., angle ~ quina) %>% 
  apa_print %>% 
  .$statistic %>% 
  str_remove_all("\\$|\\{|\\}|\\\\|mathit")

compare_qui_non_qui_angle_plot <-
  ggplot(df_angle,
         aes(y = angle,
             x = quina)) +
  geom_boxplot() +
  geom_quasirandom(alpha = point_alpha) +
  theme_bw() +
  ylab("Edge angle (\u00B0)") +
  xlab("") +
  ggtitle(paste0("Flake Edge Angle: ", t_test_str_q_angle) )

```

Do the Quina tools have different thickness?

```{r}
df_thick <- melt(flakes[, c("number", "Thickness_25%max", "Thickness_50%max", "Thickness_75%max")])

df_thick$quina <- ifelse(df_thick$number %in% quina_id, "Quina", "Non-Quina")

df_thick$label <- substring(df_thick$variable, 11, 13)

df_thick_tt_test <- 
df_thick %>% 
  nest_by(label) %>% 
  mutate(t_test = list(
    t.test(data = data, value ~ quina) %>% 
      apa_print %>% 
      .$statistic %>% 
  str_remove_all("\\$|\\{|\\}|\\\\|mathit")
  ))

compare_qui_non_qui_thick_plot <-
  ggplot(df_thick,
         aes(y = value,
             x = quina)) +
  geom_boxplot() +
  geom_quasirandom(alpha = 0.05) +
  geom_text(data = df_thick_tt_test,
            aes(label = t_test,
                x = 1.5, 
                y = 60)) +
  theme_bw() +
  facet_wrap(~label) +
  ylab("Thickness (mm)") + 
  xlab("") 

```

Do the Quina tools have different GIUR?

```{r}
# plot GIUR on Quina  --------------------------------------------------

## calculate GIUR

# get tn
df_giur_tn <- retouch[, c(1,8,10,12,14,16,18,20,22)]

# divide tn by Tn
df_giur <- df_giur_tn[, -1] / retouch[, c(9,11,13,15,17,19,21,23)]

# compute GIUR as average of all zones
df_giur_values <- 
df_giur %>% 
  as_tibble %>% 
  rowwise %>% 
  mutate(GIUR = mean(c_across(section_1_t:section_8_t), 
                     na.rm = TRUE)) %>% 
  select(GIUR)

df_giur_values$number <- df_giur_tn$number

df_giur_clustered <- 
  left_join(df_giur_values, 
            flakes_clustered[, c("number", "cluster")]) %>% 
  drop_na() %>% 
  filter(GIUR <= 1)

df_giur_clustered$quina <- ifelse(df_giur_clustered$number %in% quina_id, "Quina", "Non-Quina")

df_giur_clustered_tt_test <- 
df_giur_clustered %>% 
    t.test(data = ., 
           GIUR ~ quina) %>% 
      apa_print %>% 
      .$statistic %>% 
  str_remove_all("\\$|\\{|\\}|\\\\|mathit")

# code for raincloud plot
source("https://gist.githubusercontent.com/benmarwick/2a1bb0133ff568cbe28d/raw/fb53bd97121f7f9ce947837ef1a4c65a73bffb3f/geom_flat_violin.R")

compare_qui_non_qui_giur_plot <-
  ggplot(df_giur_clustered) +
  aes(y = GIUR,
      x = quina) +
  geom_flat_violin(position = position_nudge(x = .25, 
                                             y = 0), 
                   alpha = .8,
                   adjust = 0.35)  +
  geom_quasirandom( 
             size = .75, 
             width = 0.25,
             alpha = point_alpha) +
  geom_boxplot(
        width = .25, 
               outlier.shape = NA, 
               alpha = 0.25
               ) +
  ggtitle(paste0("Flake GIUR: ", df_giur_clustered_tt_test)) +
  coord_flip() +
  theme_bw() +
  ylab("GIUR") +
  xlab("")

```

Plot combining multiple panels

```{r}
## plot figures together.
library(cowplot)
figure_leva_quina <-
  cowplot::plot_grid(plot_grid(compare_lev_non_lev_mass_plot, 
                               compare_lev_non_lev_maxdim_plot, 
                               compare_lev_non_lev_thick_plot ,
                               labels = "auto",
                               label_size = 12,
                               ncol = 3),
                     plot_grid( 
                               compare_qui_non_qui_thick_plot ,
                               labels = c("d"),
                               label_size = 12,
                               ncol = 1,
                               rel_widths = c(1,2)),
                     plot_grid(compare_qui_non_qui_angle_plot,
                               compare_qui_non_qui_giur_plot , 
                               labels= c("e", "f"),
                               ncol = 2,
                               #rel_widths = c(1.5, 1),
                               label_size = 12),
                     nrow = 3,
                     align = 'v',
                     axis = 'lr',
                     vjust = 1.5,
                     hjust = -0.5,
                     scale = 0.9) #

save_plot(here::here("analysis/figures/fig-1-Figures-for-Leva-and-quina.png"),
                     figure_leva_quina,
          base_aspect_ratio = 1,
          base_height = 10, #18,
          base_width =  14) #22)
```



<!-- The following line inserts a page break  -->

\newpage

# References 

<!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->

<div id="refs"></div>

\newpage

### Colophon

This report was generated on `r Sys.time()` using the following computational environment and dependencies: 

```{r colophon, cache = FALSE}
# which R packages and versions?
if ("devtools" %in% installed.packages()) devtools::session_info()
```

The current Git commit details are:

```{r}
# what commit is this file at? 
if ("git2r" %in% installed.packages() & git2r::in_repository(path = ".")) git2r::repository(here::here())  
```
